Parallel Adaptive Importance Sampling

### 2016-06-02 PAIS.tex
Updated sections 3+4

Importance sampling can be a very efficient method for sampling from a
probability distribution. A proposal density is chosen, from which we
can draw samples. Each sample is assigned a weight given by the
ratio of the target density and the proposal density at that
point. They are efficient when the proposal density is concentrated in
similar areas to the target density, and incredibly inefficient when
this is not the case. The aim of the PAIS is to use the states of a
set of parallel MCMC chains to construct a proposal distribution which
will be as close as possible to the target density. Given enough
processors, the states of all of these chains at one point
in time may be reasonably representative sample of the target density.

The proposal distribution could be constructed in many different ways,
but we choose to use a mixture distribution, made up of MCMC proposals
(in this paper, specifically the pCNL proposal from Section \ref{Sec:pCNL}). Once the proposal is
constructed, we can sample a new set of states from the proposal
distribution, and each is assigned a weight given by the ratio of the target
density and the proposal mixture distribution density. Assuming that
our proposal distribution is a good one, then the variance of the
weights will be small, and we will have many useful samples. Finally, we
need to create a set of evenly weighted samples which best represent
this set of weighted samples. This is achieved by implementing the
ETPF algorithm. Once we once again have a set of evenly weighted
samples that we believe represents the target distribution well, we
can iterate the process once again. The algorithm is given in more detail in Table \ref{tab:PAIS}.

We wish to sample states $x \in X$ from a posterior
probability distribution $\mu_Y$. Since we have $M$ processes, we
represent the current state of all of the Markov chains as a vector
$\X = [x_1,x_2,\ldots,x_M]^T$. We are also given a transition kernel
$\nu(\cdot,\cdot)$, which might come from an MCMC method, for example
the pCNL proposal presented in Section \ref{Sec:pCNL}.

\begin{table}
\begin{mdframed}
\begin{algorithmic}
\STATE $\X^{(0)} = \X_0 = [x_1^{(0)},x_2^{(0)},\ldots,x_M^{(0)}]^T$
\FOR{$i=0,1,2, ..., N$}
\STATE $\Y^{(i)} = [y_1^{(i)},y_2^{(i)},\ldots,y_M^{(i)}]^T, \quad y_j^{(i)} \sim
\nu(\cdot;x_j^{(i)})$
\STATE $\chi(y;\X^{(i)}) = \frac{1}{M}
\sum_{j=1}^M \nu(y;x_j^{(i)})$.
\STATE $\W^{(i)} = [w_1^{(i)},w_2^{(i)},\ldots,w_M^{(i)}]^T,$ \quad $w^{(i)}_j =
\frac{\pi(y_j^{(i)})}{\chi(y_j^{(i)};\X^{(i)})}$.
\STATE Resample using ETPF: $(\W^{(i)},\Y^{(i)}) \rightarrow (\frac{1}{M}\mathbf{1}, \X^{(i+1)})$
\ENDFOR 
\end{algorithmic}
\end{mdframed}

@@

Since the resampling does not give us a statistically identical sample
to that which is inputted, we cannot assume that the samples $\X^{(i)}$
are samples from the posterior. Therefore, as with serial
importance samplers, the weighted samples
$(\W^{(i)},\Y^{(i)})_{i=1}^N$, or their statistics, are stored.

The key here is to choose a suitable transition kernel $\nu$ such that
if $X^{(i)}$ is a decent representative sample of the posterior,
the mixture density $\chi(\cdot;\X^{(i)})$ is a reasonable
approximation of the posterior distribution. If this is the case,
the newly proposed states $\Y^{(i)}$ will also be a good (and relatively
independent) sample of the posterior with low variance in the
weights $\W^{(i)}$.

In Section \ref{Sec:Num}, we will demonstrate how the algorithm
performs, using pCNL algorithm. We do not claim that this choice is optimal, but
is simply chosen as an example to show that sharing information across
processes can improve on the original MCMC algorithm and lead to
convergence in fewer evaluations of $\G$. This is important since if
the inverse problem being tackled involves computing
the likelihood from a very large data set this could lead to a
large saving of computational cost.

@@

Efficient selection of scaling parameters in MCMC algorithms is critical to achieving optimal mixing rates and hence achieving fast convergence to the target density. One significant difficulty is finding an appropriate $\nu$ such that
$\chi$ is a close approximation to the posterior density $\pi$. If the proposal distribution is widely over-dispersed, then the algorithm will propose states a long way from the current state resulting in a relatively low acceptance rate, impacting the quality of inferences on the posterior. Similarly, if the posterior is under-dispersed, the process will take a long time to fully explore the parameter space and so the mixing rate, and hence convergence rate, will be slow. It is therefore necessary to find a proposal distribution which is slightly over-dispersed to ensure the entire posterior is explored\cite{gelman1992inference}, but is as close to the posterior as possible.

Proposal distributions which are slightly over-dispersed as described
above, can be found by tuning the variance of the proposal
distribution $\nu$ during the burn-in phase of the
algorithm. Algorithms which use this method to find optimal proposal
distributions are known as adaptive MCMC algorithms. Adaptive MH
algorithms will converge to the stationary distribution $\pi(\cdot)$,
irrespective of whether the adaptive parameter itself converges to an
optimal value, under the following
conditions\cite{roberts2007coupling,roberts2009examples}: 
\begin{enumerate}
\item The adaptive parameter exhibits diminishing adaptation, which says that the amount of movement in the adaptive parameter decreases with the length of the chain. 
\item The smallest number of steps for which the kernel has sufficiently converged, from an initial state $x$, is finite, by bounded convergence.
\end{enumerate}

Algorithms which use mixture proposals, e.g. PAIS, must tune the variance of the individual kernels within the proposal mixture. Adaptively choosing the variance of the kernels starting with a
large initial guess allows us to first explore the state space
globally, searching for the locations of the modes. Reducing the proposal variances to an
optimal value then allows us to explore each region efficiently. The
fact that we have multiple chains allows us to assess quickly and
effectively what the optimal variance of the proposal distributions
should be. This is discussed in Section~\ref{sec:adapt}.

@@

MCMC algorithms can be assessed by comparing their approximation of the posterior to the analytic distribution. To assess this, a distance metric must be chosen. Examples are the relative error between the sample moments and the true moments, or the relative L2 error between the true density, $\pi(x|D)$, and the constructed histogram. The relative error in the $m$-th moment is
\[
	\left|\frac{N^{-1}\sum_{i=1}^N \! x_i^m - \mathbb{E}[X^m]}{\mathbb{E}[X^m]}\right|,
\]
where $\{x_i\}_{i=1}^N$ is a sample of size $N$. The relative L2 error between a continuous function to a piecewise constant function, $e$, is
\begin{equation}\label{eqn:L2_error}
	e^2 = \sum\limits_{i=1}^{n_b}\left[\displaystyle\int_{R_i} \! \pi(a|D) \, \mbox{d}a - vB_i\right]^2 \Big/ \sum\limits_{i=1}^{n_b}\left[\displaystyle\int_{R_i} \! \pi(a|D) \, \mbox{d}a\right]^2,
\end{equation}
where the regions $\{R_i\}_{i=1}^{n_b}$ are the $d$-dimensional histogram bins, so that $\bigcup_i R_i \subseteq X$ and $R_i\cap R_j=\emptyset$, $n_b$ is the number of bins, $v$ is the volume of each bin, and $B_i$ is the value of the $i$th bin.

These statistics cannot be used in practise to find optimal values of
$\beta$ since they require knowledge of the analytic solution, and
the algorithm must be run for a long time to build up a
sufficiently large sample. However they can be used to assess the
ability of other indicators to find the optimal proposal
variances in a controlled setting.

The following statistics can be used specifically for importance sampling algorithms.

@@

Importance samplers assign a weight to each sample they produce based on a ratio of the posterior to the proposal at that point. This method is most efficient when the target is proportional to the proposal distribution, in this case the weights are constant and so the variance of the weights, $var(w(y))$, is zero. Hence, we would like to choose the value of $\beta$ which minimises the variance of the weights,
\[
	\beta^*_{\text{var}} = \arg\min_{\beta} \text{var}(w(y)).
\]
The mean of the estimator $\text{var}(w(y))$ is smooth enough that it can be used to tune $\beta$ during the burn-in phase of MCMC algorithms. However the variance of the estimator can be large, especially far away from the optimal value, so it can take a large number of iterations to calculate descent directions.

@@

The statistic $\neff$ is better scaled than the variance of the weights, and is easier to interpret, so is preferable as a means of tuning the scaling parameter. In the following numerics, we use the effective sample size calculated at each iteration and averaged across the entire run to tune the scaling parameters. We do this because this is the method we use to tune the scaling parameters on the fly, comparison to the optimal scaling parameter found using the global optimum of the variance of the weights is also included for comparison.

The effective sample size also has another useful property; if we imagine the algorithm in the burn-in phase, for example, we have $M$ processes in the tail of a Gaussian curve searching for the area of high density. If the processes are evenly spaced, then the particle closest to the mean will have an exponentially higher weight assigned to it. The effective sample size in this scenario will be close to 1. As the algorithm burns in, the processes find the flatter area near the mean so the number of samples contributing information to the posterior will increase. By this argument we can see that rising $\neff$ signals the end of the burn-in period.

\subsubsection{Behaviour of the effective sample size for varying ensemble size}

@@

If we are to use the effective sample size ratio as an indicator of
how to tune the proposal variance, we need to look at how it behaves
in different situations. Here we look at how the statistic behaves as we vary the ensemble size, $M$. Figure~\ref{fig:neff-M} shows results for the Gaussian posterior discussed in Section~\ref{sec:problem 1} when using the PAIS-RW algorithm. Figure~\ref{fig:neff-M} (a) shows that as the ensemble size increases, the scaling parameter which gives the optimal effective sample size ratio decreases. This is to be expected since we are trying to fit more kernels into the posterior distribution, so the variance of each kernel needs to be decreased.

Figure~\ref{fig:neff-M} (b) shows that as the ensemble size increases, the efficiency of the sampler also increases.

\subsection{The Adaptive Algorithm}\label{sec:adapt}

A popular approach for adaptive MCMC algorithms is to view the scaling
parameter as a random variable which we can sample during the course
of the MCMC iterations. However, it can be slow to converge to the
optimal value, and we may need an
uninformative prior for the scaling parameter. Alternatively, the parameter may be randomly sampled from some interval at various points during the chain. This results in some iterations which explore the state space, and some which make local moves. Algorithms of this type do not converges to an optimal value of the scaling parameter.

We choose to use a divide and conquer scheme which optimises the effective sample size (or any other diagnostic). Some more sophisticated examples are described in \cite{roberts2009examples} and \cite{Ji2013adaptive}. Algorithm~\ref{tab:adapt} in Appendix~\ref{Sec:App}, is not presented as an optimal strategy but as a demonstration that the PAIS algorithm can be adaptively tuned effectively.

From here on in, we will use the random walk proposal,
\begin{equation}Y = X_{i-1}+\beta W_{i-1}, \qquad W_{i-1} \sim \mathcal{N}(0,1),
\end{equation}
where $X_{i-1}$ is our current state and $Y$ is our proposed state. In the numerics section we will refer to $\beta$ as the scaling parameter.

 In Algorithm~\ref{tab:adapt}, we calculate a sequence $\{\beta^{(k)}\}_{k=1}$ which converges roughly to the optimal scaling parameter $\beta^*$, resulting in the optimal transition density $\chi$ for our MCMC algorithm. We must choose some sequence of iterations, $\{n_k\}_{k=1}$, at which to update $\beta$. These $n_k$ can be constructed by a sequence in which the terms grow exponentially further apart. The algorithm is given for the PAIS method but a similar method is used to adaptively calculate the optimal scaling parameter, $\beta^*$, for the MH algorithm.

When implemented, there are a number of parameters which must be
chosen to ensure efficient tuning of $\beta$. Firstly, the initial
value of $\beta$ should be chosen so that the chains quickly explore the parameter space. However, if $\beta$ is chosen to be too
large, then the method may become unstable. We also choose two
iteration numbers at which we change our adaptive strategy. At
iteration $N_\text{join}$ we resample using the full ensemble instead
of using two subsets of the full ensemble. When using two subsets we
are training the scaling parameter using the wrong ensemble size, this
allows faster convergence but leads to a sub-optimal scaling
parameter. Resampling with the full ensemble towards the end of the
tuning allows fast convergence to the right order of magnitude,
followed by fine tuning to the optimal scaling parameter. Finally we
choose a time, $N_\text{stop}$, at which we stop updating our
parameter.

@@

In this example we compare the naively parallelised RWMH algorithm with its
PAIS variant, the PAIS-RW algorithm. We perform this comparison using the relative L2 error defined in Equation~\ref{eqn:L2_error}, as well as the relative error in the first moment.

@@

Consider the simple case of a linear observation operator $\G(u) = u$, where the prior on $u$ and the observational noise follow Gaussian distributions. Then, following Equation~\ref{eqn:like}, the Gaussian posterior has the form
\begin{equation}\label{eqn:Gaussian posterior}
	\text{law}(\mu_Y) = \pi(u|D) \propto \exp\left(-\frac{1}{2}\big\|u - D\big\|^2_{\sigma^2} - \frac{1}{2}\big\|u\big\|^2_{\tau^2}\right),
\end{equation}
where $\sigma^2$ and $\tau^2$ are the variances of the
observational noise and prior distributions respectively. In the
numerics which follow, we choose $\tau^2 =0.01$ and $\sigma^2 = 0.01$, and
we observe $u=4$ noisily such that
\[
D = \mathcal{G}(u) + \eta \sim \mathcal{N}(\mathcal{G}(u),\sigma^2).
\]
These values result in a posterior density in which the vast majority
of the density is out in the tails of the
prior distribution. The Kullback-Leibler (KL)
divergence, which gives us a measure of how different the prior and
posterior are, is $D_{KL}(\mu_Y || \mu_0) = 4.67$ for this problem.

@@

{\bf (2) Measuring convergence of nonadaptive algorithms}: We run the algorithms in Section~\ref{sec:problem 1} for one million iterations, again with $M=50$. The simulations are repeated 32 times using the optimal parameters found in (1). The performance of the algorithms is judged by the convergence of the relative L2 error statistic in Equation~\ref{eqn:L2_error}.

{\bf (3) Measuring convergence of adaptive algorithms}: We run the
adaptive algorithms under the same conditions as the nonadaptive
algorithms, and again use the relative L2 error to compare
efficiency. The adaptive algorithms are initialised with $\beta^{(1)}$ = 2.

@@

\begin{table}[!htb]
    \begin{minipage}{.5\linewidth}
      \centering
        \begin{tabular}{|c|r|}
	\hline
	Statistic											& RWMH \\ \hline
	$\beta_{\text{L2}}^*$								& 2.1e-2 \\
	$\beta_{\%}^*$									& 1.5e-1 \\
	Acceptance Rate ($\beta_{\text{L2}}^*$)				& 9.0e-1 \\
	Acceptance Rate ($\beta_{\%}^*$)					& 5.0e-1 \\
	\hline
	\end{tabular}
    \end{minipage}%
    \begin{minipage}{.5\linewidth}
      \centering
        \begin{tabular}{|l|r|r|}
	\hline
	Statistic							& PAIS-RW \\ \hline
	$\beta_{\text{eff}}^*$				& 4.7e-2 \\
	$\beta_{\text{var}(w(y))}^*$		& 5.8e-2 \\
	$\beta_{\text{L2}}^*$				& 3.9e-2 \\
	\hline
	\end{tabular}
    \end{minipage}
	\vspace{1mm}
	\caption{Optimal values of $\beta$ summarised from Figure~\ref{fig:P1 opt beta}. Statistics calculated as described in Section~\ref{sec:statistics}. The values $\beta^*_{\text{L2}}$ and $\beta^*_{\%}$ are the optimal scaling parameters found by optimising the relative L2 errors and acceptance rate respectively. Similarly $\beta_{\text{eff}}^*$ and $\beta_{\text{var}(w(y))}^*$ optimise the effective sample size and variance of the weights statistics.}
	\label{table:prob1 opt beta}
\end{table}

Figure~\ref{fig:P1 opt beta} (a) shows the two values of $\beta$ which are optimal according to the acceptance rate and relative L2 error criteria for the RWMH algorithm. The smaller estimate comes from the relative L2 error, and the larger from the acceptance rate. The results in Figure~\ref{fig:P1 opt beta} are summarised in Table~\ref{table:prob1 opt beta}. Since in general we cannot calculate the relative L2 error, we must optimise the algorithm using the acceptance rate. From the relative L2 error curve we can see that the minimum is very wide and despite the optimal values being very different there is not a large difference in the convergence rate.

Figure~\ref{fig:P1 opt beta} (b) shows the effective sample size ratio compared to the error analysis and the variance of the weights. The relative L2 error graph is noisy, but it is clear that the maximum in the effective sample size and the minimum in the variance of the weights are both close to the minimum in the relative L2 error. Due to this we say that the estimate of the effective sample size found by averaging the statistic over each iteration is a good indicator for the optimal scaling parameter. In general this indicator underestimates the estimate found by using the entire sample.

@@

We can achieve these two regimes in RWMH by tuning $\beta$ using the acceptance rate for local convergence, and by L2 error for global convergence. Similarly in PAIS-RW we can use the effective sample size for local convergence, and the L2 error for global convergence.

>>>>

Importance sampling can be a very efficient method for sampling from a
probability distribution. A proposal density is chosen, from which we
can draw samples. Each sample is assigned a weight given by the
ratio of the target density and the proposal density at that
point. They are efficient when the proposal density is concentrated in
similar areas to the target density, and incredibly inefficient when
this is not the case. The aim of the PAIS is to use the states of a
ensemble ofl MCMC chains to construct a proposal distribution which
will be as close as possible to the target density. Given a big enough
ensemble, the states of all of these chains at one point
in time may be reasonably representative sample of the target density.

The proposal distribution could be constructed in many different ways,
but we choose to use a mixture distribution, made up of a sum of MCMC
proposal distributions for each of the members of the ensemble. Once the proposal is
constructed, we can sample a new set of states from the proposal
distribution, and each is assigned a weight given by the ratio of the target
density and the proposal mixture distribution density. Assuming that
our proposal distribution is a good one, then the variance of the
weights will be small, and we will have many useful samples. Finally, we
need to create a set of evenly weighted samples which best represent
this set of weighted samples. This is achieved by implementing a
resampling algorithm. Initially we will use the
ETPF algorithm\cite{reich2013nonparametric}, although we will suggest an alternative strategy in
Section \ref{sec:AMR}. The output of the resampling algorithm gives us
a set of evenly weighted
samples that we believe represents the target distribution well, and
from this point we
can iterate the process once again. The algorithm is summarised in Table \ref{tab:PAIS}.

We wish to sample states $x \in X$ from a posterior
probability distribution $\mu_D$, where $D$ represents our data which
we wish to use to infer $x$. Since we have $M$ processes, we
represent the current state of all of the Markov chains as a vector
$\X = [x_1,x_2,\ldots,x_M]^T$. We are also given a transition kernel
$\nu(\cdot,\cdot)$, which might come from an MCMC method, for example
the random walk Metropolis-Hastings proposal density $\nu(\cdot,x) ~
\mathcal{N}(x,\beta^2)$, where $\beta^2\in \mathbb{R}$ defines the
variance of the proposal.

\begin{table}
\begin{mdframed}
\begin{algorithmic}
\STATE $\X^{(0)} = \X_0 = [x_1^{(0)},x_2^{(0)},\ldots,x_M^{(0)}]^T$
\FOR{$i=0,1,2, ..., N$}
\STATE $\Y^{(i)} = [y_1^{(i)},y_2^{(i)},\ldots,y_M^{(i)}]^T, \quad y_j^{(i)} \sim
\nu(\cdot;x_j^{(i)})$
\STATE $\chi(y;\X^{(i)}) = \frac{1}{M}
\sum_{j=1}^M \nu(y;x_j^{(i)})$.
\STATE $\W^{(i)} = [w_1^{(i)},w_2^{(i)},\ldots,w_M^{(i)}]^T,$ \quad $w^{(i)}_j =
\frac{\pi(y_j^{(i)})}{\chi(y_j^{(i)};\X^{(i)})}$.
\STATE Resample: $(\W^{(i)},\Y^{(i)}) \rightarrow (\frac{1}{M}\mathbf{1}, \X^{(i+1)})$
\ENDFOR 
\end{algorithmic}
\end{mdframed}

@@

Since the resampling does not give us a statistically identical sample
to that which is inputted, we cannot assume that the samples $\X^{(i)}$
are samples from the posterior. Therefore, as with serial
importance samplers, the weighted samples
$(\W^{(i)},\Y^{(i)})_{i=1}^N$ are the samples from the posterior that
we will analyse.

The key is to choose a suitable transition kernel $\nu$ such that
if $X^{(i)}$ is a good representative sample of the posterior,
then the mixture density $\chi(\cdot;\X^{(i)})$ is a good
approximation of the posterior distribution. If this is the case,
the newly proposed states $\Y^{(i)}$ will also be a good sample of the posterior with low variance in the
weights $\W^{(i)}$.

In Section \ref{Sec:Num}, we will demonstrate how the algorithm
performs, primarily using RWMH proposals. We do not claim that this choice is optimal, but
is simply chosen as an example to show that sharing information across
processes can improve on the original MCMC algorithm and lead to
convergence in fewer evaluations of $\G$. This is important since if
the inverse problem being tackled involves computing
the likelihood from a very large data set this could lead to a
large saving of computational cost. We have observed that
using more complex (and hence more expensive) kernels $\nu$, does not
significantly improve the speed of convergence of the algorithm.

Care needs to be taken when choosing the proposal distribution to
ensure that the proposals are absolutely continuous with respect to
the posterior distribution. In Section \ref{sec:chem} we will consider
an inverse problem with Gamma priors. Since the posterior distribution
in this context is heavy-tailed, if we use a lighter tailed
distribution for $\nu$, such as a Gaussian kernel, then importance
weights in the tails will be unbounded, which will hamper convergence
of the algorithm.

@@

Efficient selection of scaling parameters in MCMC algorithms is
critical to achieving optimal mixing rates and hence achieving fast
convergence to the target density. One aspect worthy of  consideration
with the PAIS, is finding an appropriate $\nu$ such that
$\chi$ is a close approximation to the posterior density $\pi$. If the
proposal distribution is over-dispersed, then the algorithm will often
propose states in the tails of the distribution, resulting in larger
variance of the weights, and therefore slower converegence to the
posterior distribution. Similarly, if the proposal distribution is
under-dispersed, the proposals will be highly correlated with the
previous states, and the algorithm will take a long time to fully explore the parameter spac. It is therefore necessary to find a proposal distribution which is slightly over-dispersed to ensure the entire posterior is explored\cite{gelman1992inference}, but is as close to the posterior as possible.

Most commonly used MCMC proposals have parametric dependence which
allows the user to control their variance. For example, in the RWMH
proposal $y = x + \beta \eta$, the parameter $\beta$ controls how
correlated the proposal state $y$ is to the current state
$x$. Therefore the proposal distributions can be tuned such that they are are slightly over-dispersed, as described
above. This tuning can take place during the burn-in phase of the
algorithm. Algorithms which use this method to find optimal proposal
distributions are known as adaptive MCMC algorithms, and have been
shown to be convergent provided that they satisfy certain conditions\cite{roberts2007coupling,roberts2009examples}.

Algorithms which use mixture proposals, e.g. PAIS, must tune the
variance of the individual kernels within the proposal mixture. This
adaptivity during the burn-in has some added benefits over and above
finding an optimal parameter regime for the algorithm. If the initial
value of the proposal variances is chosen to be very large, then
proposed moves will be made far and wide, expediting the early
mode-finding stages of the algorithm. Adaptively reducing the proposal variances to an
optimal value then allows us to explore each region efficiently. The
fact that we have an ensemble of chains allows us to assess quickly and
effectively what the optimal variance of the proposal distributions
should be.

@@

MCMC algorithms can be assessed by comparing their approximation of
the posterior to the analytic distribution, in cases where the
posterior distribution can be computed using alternative methods. To
assess this, a distance metric on distributions must be chosen. Examples are the
relative error between the sample moments and the posterior's moments, or the
relative $L^2$ error between the true density, $\pi(x|D)$, and the
constructed histogram. The relative error in the $m$-th moment is
given by:
\[
	\left|\frac{N^{-1}\sum_{i=1}^N \! x_i^m - \mathbb{E}[X^m]}{\mathbb{E}[X^m]}\right|,
\]
where $\{x_i\}_{i=1}^N$ is a sample of size $N$. The relative $L^2$
error between a continuous function to a piecewise constant function,
$e$, can be given by considering the difference in mass between the
normalised histogram of the samples and the posterior distribution
over a set of disjoint sets or ``bins'':
\begin{equation}\label{eqn:L2_error}
	e^2 = \sum\limits_{i=1}^{n_b}\left[\displaystyle\int_{R_i} \! \pi(a|D) \, \mbox{d}a - vB_i\right]^2 \Big/ \sum\limits_{i=1}^{n_b}\left[\displaystyle\int_{R_i} \! \pi(a|D) \, \mbox{d}a\right]^2,
\end{equation}
where the regions $\{R_i\}_{i=1}^{n_b}$ are the $d$-dimensional histogram bins, so that $\bigcup_i R_i \subseteq X$ and $R_i\cap R_j=\emptyset$, $n_b$ is the number of bins, $v$ is the volume of each bin, and $B_i$ is the value of the $i$th bin.

These statistics cannot be used in general to find optimal values of
$\beta$ since they require knowledge of the analytic solution, and
the algorithm must be run for a long time to build up a
sufficiently large sample. However they can be used to assess the
ability of other indicators to find the optimal proposal
variances in a controlled setting. The following statistics can be used specifically for importance sampling algorithms.

@@

Importance samplers assign a weight to each sample they produce based
on a ratio of the posterior to the proposal at that point. Importance
samplers are most efficient when the target is proportional to the
proposal distribution. In this case the weights are all equal, and so the variance of the weights, $var(w(y))$, is zero. Hence, we would like to choose the value of $\beta$ which minimises the variance of the weights,
\[
	\beta^*_{\text{var}} = \argmin_{\beta} \text{var}(w(y)).
\]
In our experience, the mean of the estimator
$\text{var}(w(y))$ is a smooth enough function of $\beta$ that it can
be used to tune the proposal variance during the burn-in phase of MCMC
algorithms. However the variance of the estimator of the variance can be large, especially far away from the optimal value, so it can take a large number of iterations to calculate descent directions.

@@

The statistic $\neff$ is easier to deal with than the variance of the
weights, as it varies between $1$ and $M$ as opposed to the variance
which can vary over many orders of magnitude.
Therefore it is preferable as a means of tuning the scaling
parameter. In the following numerics, we use the effective sample size
calculated at each iteration and averaged across the entire run to
tune the scaling parameters. We do this because when we tune this
parameter on the fly, we use the single-iteration average to estimate
optimality. The optimal scaling parameter found using the global
optimum of the variance of the weights is also included for
comparison. Note that for the majority of iterations the value of
$\neff$ is an overestimate of this statistic over a larger number of
samples. Rare events which result in a large importance weight bring
this statistic down, and care must be made not to overfit the proposal
variance to $\neff$ on an iteration by iteration basis. This can be
accounted for by increasing the variance slightly once the adaptive
algorithm has arrived on a value using single iteration values of $\neff$.

The effective sample size also has another useful property; if we
imagine the algorithm in the burn-in phase, for example, we have $M$
processes in the tail of a Gaussian curve searching for the area of
high density. If the processes are evenly spaced, then the particle
closest to the mean will have an exponentially higher weight assigned
to it. The effective sample size in this scenario will be close to
1. As the algorithm burns in, the ensemble populates the regions where
the majority of the probability density lies, and the proposal
distributions better represent the posterior distribution. This leads
to smaller variance in the weights, and a bigger effective sample size.
By this argument we can see that rising $\neff$ signals the end of the burn-in period.

@@

If we are to use the effective sample size ratio as an indicator of
how to tune the proposal variance, we need to look at how it behaves
in different situations. Here we look at how the statistic behaves as
we vary the ensemble size, $M$. Figure~\ref{fig:neff-M} shows results
for the Gaussian posterior discussed in Section~\ref{sec:problem 1}
when using the PAIS algorithm with RWMH proposals, each with variance
$\beta^2 \in \mathbb{R}_{>0}$. Figure~\ref{fig:neff-M} (a) shows that
as the ensemble size increases, the scaling parameter which gives the
optimal effective sample size ratio decreases. This is to be expected
since if we are trying to approximate the posterior with a mixture
distribution of a small number of Gaussians, the optimal variance will
naturally be larger so that the whole of the significant regions are
covered by the proposal distribution. As the number increases, the
optimal variances decrease so that finer details in the posterior can
be better represented in the proposal. Figure~\ref{fig:neff-M} (b) shows that as the ensemble size increases, the efficiency of the sampler also increases.

\subsection{Adaptive PAIS}\label{sec:adapt}

A popular approach for adaptive MCMC algorithms is to view the scaling
parameter as a random variable which we can sample during the course
of the MCMC iterations. However, it can be slow to converge to the
optimal value, and we may need an
uninformative prior for the scaling parameter. Alternatively, the
parameter may be randomly sampled at various points during the
evolution of the chain. This results in some iterations which make
larger global moves in the state space between modes in the target
distribution, and some which make local moves. Algorithms of this type
do not converge to an optimal value of the scaling parameter. We choose to use a divide and conquer scheme which optimises the
effective sample size (or any other diagnostic). Some more
sophisticated examples are described in \cite{roberts2009examples} and
\cite{Ji2013adaptive}. 
% Algorithm~\ref{tab:adapt} in Appendix~\ref{Sec:App}, is not presented as an optimal strategy but as a demonstration that the PAIS algorithm can be adaptively tuned effectively.

From here on in, we will use the random walk proposal,
\begin{equation}y = x_{i-1}+\beta \omega_{i-1}, \qquad \omega_{i-1}
  \sim \mathcal{N}(0,1) \qquad {\rm i.i.d.}
\end{equation}
where $x_{i-1}$ is our current state and $y$ is our proposed state. In the numerics section we will refer to $\beta$ as the scaling parameter.

 Using an adaptive strategy, we calculate a sequence
 $\{\beta^{(k)}\}_{k=1}$ which converges roughly to the optimal
 scaling parameter $\beta^*$, resulting in the optimal transition
 density $\chi$ for our MCMC algorithm. We must choose some sequence
 of iterations, $\{n_k\}_{k=1}$, at which to update $\beta$, and due
 to the constraints on adaptive MCMC
 algorithms\cite{roberts2007coupling,roberts2009example}, these $n_k$
 must grow exponentially further apart. This same adaptive approach
 can also be applied to the trivially parallelised MCMC algorithms to adaptively calculate their optimal scaling parameter $\beta^*$.

% When implemented, there are a number of parameters which must be
% chosen to ensure efficient tuning of $\beta$. Firstly, the initial
% value of $\beta$ should be chosen so that the chains quickly explore the parameter space. However, if $\beta$ is chosen to be too
% large, then the method may become unstable. We also choose two
% iteration numbers at which we change our adaptive strategy. At
% iteration $N_\text{join}$ we resample using the full ensemble instead
% of using two subsets of the full ensemble. When using two subsets we
% are training the scaling parameter using the wrong ensemble size, this
% allows faster convergence but leads to a sub-optimal scaling
% parameter. Resampling with the full ensemble towards the end of the
% tuning allows fast convergence to the right order of magnitude,
% followed by fine tuning to the optimal scaling parameter. Finally we
% choose a time, $N_\text{stop}$, at which we stop updating our
% parameter.

@@

In this example we compare the naively parallelised RWMH algorithm with its
PAIS variant, the PAIS-RW algorithm. We perform this comparison using the relative $L^2$ error defined in Equation~\ref{eqn:L2_error}, as well as the relative error in the first moment.

@@

Consider the simple case of a linear observation operator $\G(u) = u$, where the prior on $u$ and the observational noise follow Gaussian distributions. Then, following Equation~\ref{eqn:like}, the Gaussian posterior has the form
\begin{equation}\label{eqn:Gaussian posterior}
	\text{law}(\mu_D) = \pi(u|D) \propto \exp\left(-\frac{1}{2}\big\|u - D\big\|^2_{\sigma^2} - \frac{1}{2}\big\|u\big\|^2_{\tau^2}\right),
\end{equation}
where $\sigma^2$ and $\tau^2$ are the variances of the
observational noise and prior distributions respectively. In the
numerics which follow, we choose $\tau^2 =0.01$ and $\sigma^2 = 0.01$, and
we observe $u=4$ noisily such that
\[
D = \mathcal{G}(u) + \eta \sim \mathcal{N}(\mathcal{G}(u),\sigma^2).
\]
These values result in a posterior density in which the vast majority
of the density is out in the tails of the
prior distribution. The Kullback-Leibler (KL)
divergence, which gives us a measure of how different the prior and
posterior are, is $D_{KL}(\mu_D || \mu_0) = 4.67$ for this problem.

@@

{\bf (2) Measuring convergence of nonadaptive algorithms}: We run the algorithms in Section~\ref{sec:problem 1} for one million iterations, again with $M=50$. The simulations are repeated 32 times using the optimal parameters found in (1). The performance of the algorithms is judged by the convergence of the relative $L^2$ error statistic in Equation~\ref{eqn:L2_error}.

{\bf (3) Measuring convergence of adaptive algorithms}: We run the
adaptive algorithms under the same conditions as the nonadaptive
algorithms, and again use the relative $L^2$ error to compare
efficiency. The adaptive algorithms are initialised with $\beta^{(1)}$ = 2.

@@

\begin{table}[!htb]
    \begin{minipage}{.5\linewidth}
      \centering
        \begin{tabular}{|c|r|}
	\hline
	Statistic											& RWMH \\ \hline
	$\beta_{\text{L2}}^*$								& 2.1e-2 \\
	$\beta_{\%}^*$									& 1.5e-1 \\
	Acceptance Rate ($\beta_{\text{L2}}^*$)				& 9.0e-1 \\
	Acceptance Rate ($\beta_{\%}^*$)					& 5.0e-1 \\
	\hline
	\end{tabular}
    \end{minipage}%
    \begin{minipage}{.5\linewidth}
      \centering
        \begin{tabular}{|l|r|r|}
	\hline
	Statistic							& PAIS-RW \\ \hline
	$\beta_{\text{eff}}^*$				& 4.7e-2 \\
	$\beta_{\text{var}(w(y))}^*$		& 5.8e-2 \\
	$\beta_{\text{L2}}^*$				& 3.9e-2 \\
	\hline
	\end{tabular}
    \end{minipage}
	\vspace{1mm}
	\caption{Optimal values of $\beta$ summarised from Figure~\ref{fig:P1 opt beta}. Statistics calculated as described in Section~\ref{sec:statistics}. The values $\beta^*_{\text{L2}}$ and $\beta^*_{\%}$ are the optimal scaling parameters found by optimising the relative $L^2$ errors and acceptance rate respectively. Similarly $\beta_{\text{eff}}^*$ and $\beta_{\text{var}(w(y))}^*$ optimise the effective sample size and variance of the weights statistics.}
	\label{table:prob1 opt beta}
\end{table}

Figure~\ref{fig:P1 opt beta} (a) shows the two values of $\beta$ which are optimal according to the acceptance rate and relative $L^2$ error criteria for the RWMH algorithm. The smaller estimate comes from the relative $L^2$ error, and the larger from the acceptance rate. The results in Figure~\ref{fig:P1 opt beta} are summarised in Table~\ref{table:prob1 opt beta}. Since in general we cannot calculate the relative $L^2$ error, we must optimise the algorithm using the acceptance rate. From the relative $L^2$ error curve we can see that the minimum is very wide and despite the optimal values being very different there is not a large difference in the convergence rate.

Figure~\ref{fig:P1 opt beta} (b) shows the effective sample size ratio compared to the error analysis and the variance of the weights. The relative $L^2$ error graph is noisy, but it is clear that the maximum in the effective sample size and the minimum in the variance of the weights are both close to the minimum in the relative $L^2$ error. Due to this we say that the estimate of the effective sample size found by averaging the statistic over each iteration is a good indicator for the optimal scaling parameter. In general this indicator underestimates the estimate found by using the entire sample.

@@

We can achieve these two regimes in RWMH by tuning $\beta$ using the acceptance rate for local convergence, and by $L^2$ error for global convergence. Similarly in PAIS-RW we can use the effective sample size for local convergence, and the $L^2$ error for global convergence.
